---
title: "02-EMODnet-Biology-phytoplankton-alpha-diversity"
author: "Anders Torstensson, Lisa Sundqvist and Markus Lindh"
date: '2023-03-02'
params:
  width: 1280 # Output width in px
  res: 200 # Output resolution in dpi
knit: (function(inputFile, encoding) {
                        rmarkdown::render(inputFile,
                        encoding=encoding, 
                        output_file="../docs/02_maps.html") })
---

This R markdown produces monthly png maps of phytoplankton species diversity (Shannon diversity index) and richness, and plots the diversity and richness for the entire region (gamma diversity) over time together with sampling effort (n samples).

```{r setup, echo=FALSE, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(lubridate)
library(EMODnetBiologyMaps)
library(viridis)
library(gganimate)

derivedDir = "../data/derived_data"
mapsDir = "../product/maps"
plotsDir = "../product/plots"
```

## Load previous objects

Objects produced in 01_phytoplankton_alpha_diversity_netcdf.Rmd

```{r load_objects, echo=FALSE}
load(file.path(derivedDir, "spdf_data_monthly.Rda"))
load(file.path(derivedDir, "gamma_diversity.Rda"))
```

## Production of richness maps

Produce png maps of species richness for each month of the time series (derived from the animation). 

Maps are stored in ../product/maps/richness/

```{r produce_maps_richness, message=FALSE, results=FALSE, echo=FALSE}
start_time = Sys.time()

animated_map = emodnet_map_basic(crs = 4326, 
                                 xlim = c(10, 29),
                                 ylim = c(54, 66)
                                 ) +
  geom_point(data = spdf_data_monthly, 
             aes(x = decimalLongitude, y = decimalLatitude, group = as.factor(stationCluster), 
                 color = uniqueTaxa, 
                 size = uniqueTaxa)) +
  transition_time(as.integer(daycount)) +
  xlab("") +
  ylab("") +
  scale_color_viridis("Unique taxa",
                      option="plasma",
                      limits = c(min(spdf_data_monthly$uniqueTaxa, na.rm = TRUE),
                                 max(spdf_data_monthly$uniqueTaxa, na.rm = TRUE)),
                      breaks = seq(5,
                                   max(spdf_data_monthly$uniqueTaxa, na.rm = TRUE),
                                   by = 10)) +
  scale_size_continuous("Unique taxa",
                        # range = c(2,8),
                        limits = c(min(spdf_data_monthly$uniqueTaxa, na.rm = TRUE),
                                   max(spdf_data_monthly$uniqueTaxa, na.rm = TRUE)),
                        breaks = seq(5,
                                     max(spdf_data_monthly$uniqueTaxa, na.rm = TRUE),
                                     by = 10)) +
  guides(color = guide_legend(), size = guide_legend()) +
  labs(title = "Phytoplankton species richness",
       subtitle = '{format(as_date(frame_time), "%B %Y")}')

animate(
      plot = animated_map,
      device = "png",
      renderer = file_renderer(file.path(mapsDir, "richness"), prefix = "richness_", overwrite = TRUE),
      nframes = length(unique(spdf_data_monthly$monthYear)),
      width = params$width, height = 0.8*params$width, units = "px",
      res = params$res)

end_time = Sys.time()
end_time - start_time
```

```{r elapsed_time_1, echo=FALSE}
end_time = Sys.time()
end_time - start_time
```

## Production of shannon maps

Produce png maps of the shannon diversity index for each month of the time series (derived from the animation). 

Maps are stored in ../product/maps/shannon/

```{r produce_maps_shannon, message=FALSE, results=FALSE, echo=FALSE}
start_time = Sys.time()

animated_map_shannon = emodnet_map_basic(crs = 4326, 
                                         xlim = c(10, 29),
                                         ylim = c(54, 66)
                                         ) +
  geom_point(data = spdf_data_monthly, 
             aes(x = decimalLongitude, y = decimalLatitude, group = as.factor(stationCluster), 
                 color = shannon, 
                 size = shannon)) +
  transition_time(as.integer(daycount)) +
  xlab("") +
  ylab("") +
  scale_color_viridis("Shannon index",
                      option="viridis",
                      limits = c(min(spdf_data_monthly$shannon, na.rm = TRUE),
                                 max(spdf_data_monthly$shannon, na.rm = TRUE)),
                      breaks = seq(.5,
                                   max(spdf_data_monthly$shannon, na.rm = TRUE),
                                   by = 0.25)) +
  scale_size_continuous("Shannon index",
                        # range = c(2,8),
                        limits = c(min(spdf_data_monthly$shannon, na.rm = TRUE),
                                   max(spdf_data_monthly$shannon, na.rm = TRUE)),
                        breaks = seq(.5,
                                     max(spdf_data_monthly$shannon, na.rm = TRUE),
                                     by = 0.25)) +
  guides(color = guide_legend(), size = guide_legend()) +
  labs(title = "Phytoplankton biodiversity",
       subtitle = '{format(as_date(frame_time), "%B %Y")}')

animate(
      plot = animated_map_shannon,
      device = "png",
      renderer = file_renderer(file.path(mapsDir, "shannon"), prefix = "shannon_", overwrite = TRUE),
      nframes = length(unique(spdf_data_monthly$monthYear)),
      width = params$width, height = 0.8*params$width, units = "px",
      res = params$res)
```

```{r elapsed_time_2, echo=FALSE}
end_time = Sys.time()
end_time - start_time
```

## Plot gamma richness

Plot the monthly richness of the whole dataset over time, together with sampling effort.

Maps are stored in ../product/plots/

```{r plot_gamma_richness, echo=FALSE}
color_month = c("#529BB5", "#65AD97", "#92BC63", "#D0B440", "#E59036", "#E1562A", 
                "#AE221D", "#521913", "#C6AED4", "#A16CAD", "#6C4F9E", "#4E79C4")

scaleFactor_richness = max(gamma_diversity$n) / max(gamma_diversity$uniqueTaxa)


gamma_richness.p = gamma_diversity %>%
  ggplot(aes(x = date)) +
  geom_bar(aes(y = n/scaleFactor_richness, group=date, fill = month), stat = "identity", size = .1, color = "black", alpha = .5) + 
  geom_line(aes(y = uniqueTaxa), color = "blue", size = 1) +
  scale_y_continuous(
    name = "Unique taxa",
    sec.axis = sec_axis(~.*scaleFactor_richness, name="Sample size (n)")) +   
  scale_fill_gradientn(name="Month", colours = color_month,
                       breaks=c(3,6,9,12),
                       labels=c("Mar", "Jun", "Sep", "Dec")) +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5, size=14, face="bold.italic"),
        axis.title.y.left=element_text(color="blue"),
        axis.text.y.left=element_text(color="blue")) +
  xlab("Year") +
  ggtitle("Gamma richness") +
  guides(fill = guide_colorbar(reverse=T))

gamma_richness.p

ggsave(file.path(plotsDir, "gamma_richness.png"), plot = gamma_richness.p)
```

## Plot gamma diversity

Plot the monthly shannon index of the whole dataset over time, together with sampling effort.

Maps are stored in ../product/plots/

```{r plot_gamma_diversity, echo=FALSE}
scaleFactor = max(gamma_diversity$n) / max(gamma_diversity$shannon)

gamma_diversity.p = gamma_diversity %>%
  ggplot(aes(x = date)) +
  geom_bar(aes(y = n/scaleFactor, group=date, fill = month), stat = "identity", size = .1, color = "black", alpha = .5) + 
  geom_line(aes(y = shannon), color = "darkgreen", size = 1) +
  scale_y_continuous(
    name = "Shannon index",
    sec.axis = sec_axis(~.*scaleFactor, name="Sample size (n)")) + 
  scale_fill_gradientn(name="Month", colours = color_month,
                       breaks=c(3,6,9,12),
                       labels=c("Mar","Jun", "Sep", "Dec")) +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5, size=14, face="bold.italic"),
        axis.title.y.left = element_text(color="darkgreen"),
        axis.text.y.left = element_text(color="darkgreen")) +
  xlab("Year") +
  ggtitle("Gamma diversity") + 
  guides(fill = guide_colorbar(reverse=T))

gamma_diversity.p

ggsave(file.path(plotsDir, "gamma_diversity.png"), plot = gamma_diversity.p)
```

### Reproducibility

```{r reproducibility}
# Date time
Sys.time()
# Here we store the session info for this script
sessioninfo::session_info()
```
Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
