---
title: "03-EMODnet-Biology-phytoplankton-alpha-diversity"
author: "Anders Torstensson, Lisa Sundqvist and Markus Lindh"
date: '2023-03-02'
params:
  fps: 5 # Define framerate
  width: 1280 # Output width in px
  res: 200 # Output resolution in dpi
  detail: 1 # Output detail
  frame_interpolations: 4
knit: (function(inputFile, encoding) {
                        rmarkdown::render(inputFile,
                        encoding=encoding, 
                        output_file="../docs/03_animations.html") })
---

This R markdown produces animations (in GIF and MP4) of phytoplankton alpha and gamma diversity over time for the station clusters defined in 01_phytoplankton_alpha_diversity_netcdf.Rmd. Please note that the high-res rendering of animations are memory intense, and may require enough RAM for multimedia creations (the machine used for writing this script used 32 GB). The frame-resolution may be adjusted under params, if necessary.

```{r setup, echo=FALSE, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(plyr)
library(tidyverse)
library(lubridate)
library(EMODnetBiologyMaps)
library(viridis)
library(geosphere)
library(gganimate)
library(magick)
library(gifski)
library(av)

derivedDir = "../data/derived_data"
animationsDir = "../product/animations"
```

## Load previous objects

Objects produced in 01_phytoplankton_alpha_diversity_netcdf.Rmd

```{r load_objects, echo=FALSE}
load(file.path(derivedDir, "spdf_data_monthly.Rda"))
load(file.path(derivedDir, "gamma_diversity.Rda"))
```

## Production of richness maps

Map animation of changes in monthly species richness (n taxa) over time, exported as GIF.

Exported as ../product/animations/animation_richness.gif

```{r produce_maps_richness, message=FALSE, results=FALSE, echo=FALSE}
start_time = Sys.time()

animated_map = emodnet_map_basic(crs = 4326, 
                                 xlim = c(10, 29),
                                 ylim = c(54, 66)
                                 ) +
  geom_point(data = spdf_data_monthly, 
             aes(x = decimalLongitude, y = decimalLatitude, group = as.factor(stationCluster), 
                 color = uniqueTaxa, 
                 size = uniqueTaxa)) +
  transition_time(as.integer(daycount)) +
  xlab("") +
  ylab("") +
  scale_color_viridis("Unique taxa",
                      option="plasma",
                      limits = c(min(spdf_data_monthly$uniqueTaxa, na.rm = TRUE),
                                 max(spdf_data_monthly$uniqueTaxa, na.rm = TRUE)),
                      breaks = seq(5,
                                   max(spdf_data_monthly$uniqueTaxa, na.rm = TRUE),
                                   by = 10)) +
  scale_size_continuous("Unique taxa",
                        limits = c(min(spdf_data_monthly$uniqueTaxa, na.rm = TRUE),
                                   max(spdf_data_monthly$uniqueTaxa, na.rm = TRUE)),
                        breaks = seq(5,
                                     max(spdf_data_monthly$uniqueTaxa, na.rm = TRUE),
                                     by = 10)) +
  guides(color = guide_legend(), size = guide_legend()) +
  labs(title = "Phytoplankton species richness",
       subtitle = '{format(as_date(frame_time), "%Y-%m")}')

animated_richness = animate(
      plot = animated_map, 
      renderer = gifski_renderer(),
      nframes = length(unique(spdf_data_monthly$monthYear))*params$frame_interpolations,
      width = params$width, height = params$width*0.8, units = "px",
      res = params$res,
      fps = params$fps,
      detail = params$detail,
      end_pause = params$fps*10)

main_gif = image_read(animated_richness)

logo = "logo_col_no_bg.png"
logo = system.file("extdata", logo, package = "EMODnetBiologyMaps", mustWork = TRUE)
logo = image_read(logo)
logo = image_scale(logo, paste(0.09375*params$width))

plot_logo_richness = image_composite(main_gif, logo, gravity = "southeast", offset = paste0("+",round(0.2890625*params$width),"+",round(0.08789062*0.8*params$width)))

image_write(plot_logo_richness, file.path(animationsDir, "animation_richness.gif"))

gc() # Not pretty, but speeds up rendering
```

```{r elapsed_time_1, echo=FALSE}
end_time = Sys.time()
end_time - start_time
```

## Production of shannon maps

Map animation of changes in monthly alpha diversity (Shannon diversity index) over time, exported as GIF.

Exported as ../product/animations/animation_shannon.gif

```{r produce_maps_shannon, message=FALSE, results=FALSE, echo=FALSE}
start_time = Sys.time()

animated_map_shannon = emodnet_map_basic(crs = 4326, 
                                         xlim = c(10, 29),
                                         ylim = c(54, 66)
                                         ) +
  geom_point(data = spdf_data_monthly, 
             aes(x = decimalLongitude, y = decimalLatitude, group = as.factor(stationCluster), 
                 color = shannon, 
                 size = shannon)) +
  transition_time(daycount) +
  xlab("") +
  ylab("") +
  scale_color_viridis("Shannon index",
                      option="viridis",
                      limits = c(min(spdf_data_monthly$shannon, na.rm = TRUE),
                                 max(spdf_data_monthly$shannon, na.rm = TRUE)),
                      breaks = seq(.5,
                                   max(spdf_data_monthly$shannon, na.rm = TRUE),
                                   by = 0.25)) +
  scale_size_continuous("Shannon index",
                        limits = c(min(spdf_data_monthly$shannon, na.rm = TRUE),
                                   max(spdf_data_monthly$shannon, na.rm = TRUE)),
                        breaks = seq(.5,
                                     max(spdf_data_monthly$shannon, na.rm = TRUE),
                                     by = 0.25)) +
  guides(color = guide_legend(), size = guide_legend()) +
  labs(title = "Phytoplankton biodiversity",
       subtitle = '{format(as_date(frame_time), "%Y-%m")}')

animated_shannon = animate(
      plot = animated_map_shannon, 
      renderer = gifski_renderer(),
      nframes = length(unique(spdf_data_monthly$monthYear))*params$frame_interpolations,
      width = params$width, height = params$width*0.8, units = "px",
      res = params$res,
      fps = params$fps,
      detail = params$detail,
      end_pause = params$fps*10)

main_gif = image_read(animated_shannon)

logo = "logo_col_no_bg.png"
logo = system.file("extdata", logo, package = "EMODnetBiologyMaps", mustWork = TRUE)
logo = image_read(logo)
logo = image_scale(logo, paste(0.09375*params$width))

plot_logo_shannon = image_composite(main_gif, logo, gravity = "southeast", offset = paste0("+", round(0.3046875*params$width),"+", round(0.08789062*0.8*params$width)))

image_write(plot_logo_shannon, file.path(animationsDir, "animation_shannon.gif"))

gc() # Not pretty, but speeds up rendering
```

```{r elapsed_time_2, echo=FALSE}
end_time = Sys.time()
end_time - start_time
```

## Plot gamma richness

Plot animation of changes in monthly gamma (whole region) species richness (n taxa) over time, exported as GIF.

Exported as ../product/animations/animation_gamma_richness.gif

```{r plot_gamma_richness, echo=FALSE}
start_time = Sys.time()

color_month = c("#529BB5", "#65AD97", "#92BC63", "#D0B440", "#E59036", "#E1562A", 
                "#AE221D", "#521913", "#C6AED4", "#A16CAD", "#6C4F9E", "#4E79C4")

scaleFactor_richness = max(gamma_diversity$n) / max(gamma_diversity$uniqueTaxa)

animated_gamma_richness = gamma_diversity %>%
  ggplot(aes(x = date)) +
  geom_bar(aes(y = n/scaleFactor_richness, group=date, fill = month), stat = "identity", size = .1, color = "black", alpha = .5) + 
  geom_line(aes(y = uniqueTaxa), color = "blue", linewidth = 1) +
  scale_y_continuous(
    name = "Unique taxa",
    sec.axis = sec_axis(~.*scaleFactor_richness, name="Sample size (n)")) + 
  scale_fill_gradientn(name="Month", colours = color_month,
                       breaks=c(3,6,9,12),
                       labels=c("Mar","Jun", "Sep", "Dec")) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size=14, face="bold.italic"),
        axis.title.y.left=element_text(color="blue"),
        axis.text.y.left=element_text(color="blue")) +
  xlab("Year") +
  ggtitle("Gamma richness") +
  guides(fill = guide_colorbar(reverse=T)) +
  transition_reveal(daycount) 

richness_gamma_rend = animate(
      plot = animated_gamma_richness, 
      renderer = gifski_renderer(),
      nframes = length(unique(spdf_data_monthly$monthYear))*params$frame_interpolations,
      width = params$width, height = params$width*0.4, units = "px",
      res = params$res,
      fps = params$fps,
      detail = params$detail,
      end_pause = params$fps*10)

richness_gamma = image_read(richness_gamma_rend)

image_write(richness_gamma, file.path(animationsDir, "animation_gamma_richness.gif"))

gc() # Not pretty, but speeds up rendering
```

```{r elapsed_time_3, echo=FALSE}
end_time = Sys.time()
end_time - start_time
```

## Plot gamma diversity

Plot animation of changes in monthly gamma (whole region) diversity (Shannon diversity index) over time, exported as GIF.

Exported as ../product/animations/animation_gamma_diversity.gif

```{r plot_gamma_diversity, echo=FALSE}
start_time = Sys.time()

scaleFactor_diversity = max(gamma_diversity$n) / max(gamma_diversity$shannon)

animated_gamma_diversity = gamma_diversity %>%
  ggplot(aes(x = date)) +
  geom_bar(aes(y = n/scaleFactor_diversity, group=date, fill = month), stat = "identity", size = .1, color = "black", alpha = .5) + 
  geom_line(aes(y = shannon), color = "darkgreen", linewidth = 1) +
  scale_y_continuous(
    name = "Shannon index",
    sec.axis = sec_axis(~.*scaleFactor_diversity, name="Sample size (n)")) + 
  scale_fill_gradientn(name="Month", colours = color_month,
                       breaks=c(3,6,9,12),
                       labels=c("Mar","Jun", "Sep", "Dec")) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size=14, face="bold.italic"),
        axis.title.y.left = element_text(color="darkgreen"),
        axis.text.y.left = element_text(color="darkgreen")) +
  xlab("Year") +
  ggtitle("Gamma diversity") + 
  guides(fill = guide_colorbar(reverse=T)) +
  transition_reveal(daycount)

shannon_gamma_rend = animate(
      plot = animated_gamma_diversity, 
      renderer = gifski_renderer(),
      nframes = length(unique(spdf_data_monthly$monthYear))*params$frame_interpolations,
      width = params$width, height = params$width*0.4, units = "px",
      res = params$res,
      fps = params$fps,
      detail = params$detail,
      end_pause = params$fps*10)

shannon_gamma = image_read(shannon_gamma_rend)

image_write(shannon_gamma, file.path(animationsDir, "animation_gamma_diversity.gif"))

gc() # Not pretty, but speeds up rendering
```

```{r elapsed_time_4, echo=FALSE}
end_time = Sys.time()
end_time - start_time
```

## Joint map/plot animations

Combined animation of alpha and gamma and maps and plots, exported as GIF and MP4.

Exported as:
../product/animations/animation_shannon_combined.gif
../product/animations/animation_shannon_combined.mp4
../product/animations/animation_richness_combined.gif
../product/animations/animation_richness_combined.mp4

```{r plot_maps, echo=FALSE}
start_time = Sys.time()

# Shannon only graphs

shannon_plot = image_append(c(plot_logo_shannon[1], shannon_gamma[1]), stack = TRUE)

n_frames_shannon = min(length(plot_logo_shannon), length(shannon_gamma))

for(i in 2:n_frames_shannon) {
  
  combined_x = plot_logo_shannon[i]
  combined_y = shannon_gamma[i]
  
  combined = image_append(c(combined_x, combined_y), stack = TRUE)
  
  shannon_plot = c(shannon_plot, combined)
}

image_write(shannon_plot, file.path(animationsDir, "animation_shannon_combined.gif"))

gc() # Not pretty, but speeds up rendering

image_write_video(shannon_plot, file.path(animationsDir, "animation_shannon_combined.mp4"), 
                  framerate = params$fps)

gc() # Not pretty, but speeds up rendering

image_scale(shannon_plot, "x600")

# Richness only graphs

richness_plot = image_append(c(plot_logo_richness[1], richness_gamma[1]), stack = TRUE)

n_frames_richness = min(length(plot_logo_richness), length(richness_gamma))

for(i in 2:n_frames_richness) {
  combined_x = plot_logo_richness[i]
  
  combined_y = richness_gamma[i]
  
  combined = image_append(c(combined_x, combined_y), stack = TRUE)
  
  richness_plot = c(richness_plot, combined)
}

image_write(richness_plot, file.path(animationsDir, "animation_richness_combined.gif"))

gc() # Not pretty, but speeds up rendering

image_write_video(richness_plot, file.path(animationsDir, "animation_richness_combined.mp4"), 
                  framerate = params$fps)

gc() # Not pretty, but speeds up rendering

image_scale(richness_plot, "x600")
```

```{r elapsed_time_5, echo=FALSE}
end_time = Sys.time()
end_time - start_time
```

### Reproducibility

```{r reproducibility}
# Date time
Sys.time()
# Here we store the session info for this script
sessioninfo::session_info()
```
Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
